taxize - taxonomic search and retrieval in R

### 2013-06-10 inst/doc/taxize_withcode_appB.tex

Suppose we have fuzzy coded trait table with 2 traits with 3 respectively 2 modalities:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
(traits <- \hlfunctioncall{read.table}(header = TRUE, sep = \hlstring{";"}, stringsAsFactors = FALSE, text = \hlstring{"taxon;T1M1;T1M2;T1M3;T2M1;T2M2\textbackslash{}nGammarus sp.;0;0;3;1;3\textbackslash{}nPotamopyrgus antipodarum;1;0;3;1;3\textbackslash{}nCoenagrion sp.;3;0;1;3;1\textbackslash{}nEnallagma cyathigerum;0;3;1;0;3\textbackslash{}nErythromma sp.;0;0;3;3;1\textbackslash{}nBaetis sp.;0;0;0;0;0\textbackslash{}n"}))
\end{alltt}
\begin{verbatim}
##                      taxon T1M1 T1M2 T1M3 T2M1 T2M2
## 1             Gammarus sp.    0    0    3    1    3
## 2 Potamopyrgus antipodarum    1    0    3    1    3
## 3           Coenagrion sp.    3    0    1    3    1
## 4    Enallagma cyathigerum    0    3    1    0    3
## 5           Erythromma sp.    0    0    3    3    1
## 6               Baetis sp.    0    0    0    0    0
\end{verbatim}
\end{kframe}
\end{knitrout}

And want to match this to a table with abundances:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
(abundances <- \hlfunctioncall{read.table}(header = TRUE, sep = \hlstring{";"}, stringsAsFactors = FALSE, 
    text = \hlstring{"taxon;abundance;sample\textbackslash{}nGammarus roeseli;5;1\textbackslash{}nGammarus roeseli;6;2\textbackslash{}nGammarus tigrinus;7;1\textbackslash{}nGammarus tigrinus;8;2\textbackslash{}nCoenagrionidae;10;1\textbackslash{}nCoenagrionidae;6;2\textbackslash{}nPotamopyrgus antipodarum;10;1\textbackslash{}nxxxxx;10;2\textbackslash{}n"}))
\end{alltt}
\begin{verbatim}
##                      taxon abundance sample
## 1         Gammarus roeseli         5      1
## 2         Gammarus roeseli         6      2
## 3        Gammarus tigrinus         7      1
## 4        Gammarus tigrinus         8      2
## 5           Coenagrionidae        10      1
## 6           Coenagrionidae         6      2
## 7 Potamopyrgus antipodarum        10      1
## 8                    xxxxx        10      2
\end{verbatim}
\end{kframe}
\end{knitrout}

First we do some basic data-cleaning and create a looktable, that will link taxa in trait table with taxa in the abundance table.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcomment{# first we remove ' sp.' from out trait table:}
traits$taxon_cleaned <- \hlfunctioncall{tolower}(\hlfunctioncall{gsub}(\hlstring{" sp."}, \hlstring{""}, traits$taxon))

@@

The we query the taxonomic hierarchy for both tables, since this is the backbone of this procedure:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{require}(taxize)
traits_classi <- \hlfunctioncall{classification}(\hlfunctioncall{get_uid}(traits$taxon_cleaned))
\end{alltt}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcomment{# species (as matched with trait table) by site matrix}
\hlfunctioncall{require}(reshape2)
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Loading required package: reshape2}}\begin{alltt}
\hlcomment{# reshape data to long format and name rows by samples}
L <- \hlfunctioncall{dcast}(abundances, sample ~ traits, fun.aggregate = sum, value.var = \hlstring{"abundance"})
\end{alltt}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: undefined columns selected}}\begin{alltt}
\hlfunctioncall{rownames}(L) <- L$sample
\end{alltt}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'L' not found}}\begin{alltt}
L$sample <- NULL
\end{alltt}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'L' not found}}\begin{alltt}
L
\end{alltt}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'L' not found}}\begin{alltt}

\hlcomment{# traits by species matrix}
Q <- traits[, 2:7][\hlfunctioncall{match}(\hlfunctioncall{names}(L), traits$taxon_cleaned), ]
\end{alltt}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'L' not found}}\begin{alltt}
\hlfunctioncall{rownames}(Q) <- Q$taxon_cleaned
\end{alltt}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'Q' not found}}\begin{alltt}
Q$taxon_cleaned <- NULL
\end{alltt}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'Q' not found}}\begin{alltt}
Q
\end{alltt}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'Q' not found}}\begin{alltt}

\hlcomment{# check}
\hlfunctioncall{all}(\hlfunctioncall{rownames}(Q) == \hlfunctioncall{colnames}(L))
\end{alltt}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'Q' not found}}\end{kframe}
\end{knitrout}

>>>>

Suppose we have fuzzy coded trait table with 2 traits with 3 respectively 2 modalities:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
(traits <- \hlfunctioncall{read.table}(header = TRUE, sep = \hlstring{';'}, stringsAsFactors=FALSE, text = 'taxon;T1M1;T1M2;T1M3;T2M1;T2M2
Gammarus sp.;0;0;3;1;3
Potamopyrgus antipodarum;1;0;3;1;3
Coenagrion sp.;3;0;1;3;1
Enallagma cyathigerum;0;3;1;0;3
Erythromma sp.;0;0;3;3;1
Baetis sp.;0;0;0;0;0
'))
\end{alltt}
\begin{verbatim}
##                      taxon T1M1 T1M2 T1M3 T2M1 T2M2
## 1             Gammarus sp.    0    0    3    1    3
## 2 Potamopyrgus antipodarum    1    0    3    1    3
## 3           Coenagrion sp.    3    0    1    3    1
## 4    Enallagma cyathigerum    0    3    1    0    3
## 5           Erythromma sp.    0    0    3    3    1
## 6               Baetis sp.    0    0    0    0    0
\end{verbatim}
\end{kframe}
\end{knitrout}

And want to match this to a table with abundances:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
(abundances <- \hlfunctioncall{read.table}(header = TRUE, sep = \hlstring{';'}, stringsAsFactors=FALSE, text = 'taxon;abundance;sample
Gammarus roeseli;5;1
Gammarus roeseli;6;2
Gammarus tigrinus;7;1
Gammarus tigrinus;8;2
Coenagrionidae;10;1
Coenagrionidae;6;2
Potamopyrgus antipodarum;10;1
xxxxx;10;2
'))
\end{alltt}
\begin{verbatim}
##                      taxon abundance sample
## 1         Gammarus roeseli         5      1
## 2         Gammarus roeseli         6      2
## 3        Gammarus tigrinus         7      1
## 4        Gammarus tigrinus         8      2
## 5           Coenagrionidae        10      1
## 6           Coenagrionidae         6      2
## 7 Potamopyrgus antipodarum        10      1
## 8                    xxxxx        10      2
\end{verbatim}
\end{kframe}
\end{knitrout}

First we do some basic data-cleaning and create a lookup-table, that will link taxa in trait table with the abundance table.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcomment{# first we remove ' sp.' from out trait table:}
traits$taxon_cleaned <- \hlfunctioncall{tolower}(\hlfunctioncall{gsub}(\hlstring{" sp."}, \hlstring{""}, traits$taxon))

@@

The we query the taxonomic hierarchy for both tables, this will be the backbone of this procedure:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{require}(taxize)
traits_classi <- \hlfunctioncall{classification}(\hlfunctioncall{get_uid}(traits$taxon_cleaned))
\end{alltt}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcomment{# species (as matched with trait table) by site matrix}
abundances$traits_taxa <- lookup$traits[\hlfunctioncall{match}(\hlfunctioncall{tolower}(abundances$taxon), lookup$taxon)]

\hlfunctioncall{require}(reshape2)
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Loading required package: reshape2}}\begin{alltt}
\hlcomment{# reshape data to long format and name rows by samples}
L <- \hlfunctioncall{dcast}(abundances, sample ~ traits_taxa, fun.aggregate = sum, value.var = \hlstring{"abundance"})
\hlfunctioncall{rownames}(L) <- L$sample
L$sample <- NULL
L
\end{alltt}
\begin{verbatim}
##   coenagrionidae-aggregated gammarus potamopyrgus antipodarum
## 1                        10       12                       10
## 2                         6       14                        0
\end{verbatim}
\begin{alltt}

\hlcomment{# traits by species matrix}
Q <- traits[, 2:7][\hlfunctioncall{match}(\hlfunctioncall{names}(L), traits$taxon_cleaned), ]
\hlfunctioncall{rownames}(Q) <- Q$taxon_cleaned
Q$taxon_cleaned <- NULL
Q
\end{alltt}
\begin{verbatim}
##                           T1M1 T1M2 T1M3 T2M1 T2M2
## coenagrionidae-aggregated    0    0    1    3    1
## gammarus                     0    0    3    1    3
## potamopyrgus antipodarum     1    0    3    1    3
\end{verbatim}
\begin{alltt}

\hlcomment{# check}
\hlfunctioncall{all}(\hlfunctioncall{rownames}(Q) == \hlfunctioncall{colnames}(L))
\end{alltt}
\begin{verbatim}
## [1] TRUE
\end{verbatim}
\end{kframe}
\end{knitrout}
